---
title: "Heirarchical Clustering - cleaner version"
date: "2025-10-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# START
```{r}
library(edgeR) 
library(RColorBrewer)
library(gplots)
library(biomaRt) #converting gene names
library(pheatmap)
 library(viridis)
library(dplyr)
library(tibble)
library(magrittr)
library(AnnotationDbi)
library(stats4)
library(BiocGenerics)
library(gtools)
library(openxlsx)
library(EnhancedVolcano)
library(stringr)
library(data.table)

library(DESeq2)
library(uwot)
library(gridExtra)
library(ggplot2)
library(cowplot)

library(dendsort)

# directories
resouce_dir           <- file.path(projectdirectory,"resources")
input_dir             <- file.path(projectdirectory,"input")
plot_dir              <- file.path(projectdirectory,"plots")
table_dir             <- file.path(projectdirectory,"tables")
object_dir            <- file.path(projectdirectory,"objects")

# Input 
processed_gene_counts_file  <- file.path(input_dir,"input_gene_count_matrix.csv")
metadata_tbm_file           <- file.path(input_dir, "metadata_TBM_formatted.csv")
biomart_metadata_fh         <- file.path(resouce_dir,"biomart_protein_coding_genes.tsv")
```

# --------------------

# 1. Load input data
```{r}
df   <- read.csv(processed_gene_counts_file, row.names = 1,check.names = F)

# biomart protein coding genes
gene_metadata <- read.delim(biomart_metadata_fh,stringsAsFactors = F)

# metadata
metadata           <- read.csv(metadata_tbm_file, stringsAsFactors = FALSE)
row.names(metadata) <- metadata$Sample_ID

metadata <- metadata[metadata$Sample_ID %in% names(df),]
metadata$Mortality <- as.factor(metadata$short_term_Mortality)
metadata$Gender    <- as.factor(metadata$gender)
metadata$Batch     <- as.numeric(metadata$BatchAnalysis)
metadata$cohortbatch   <- paste("TBM",metadata$Batch,sep = "_") %>% as.factor()
metadata$Status        <- metadata$death_shortterm
metadata$Age        <- metadata$age
metadata$TBM_status <- metadata$post_MNGS_TBM_status
metadata$TBM_status[metadata$TBM_status %in% "Definite TBM"] <- "DefiniteTBM"
metadata$TBM_status[metadata$TBM_status %in% "Probable TBM"] <- "ProbableTBM"
metadata$mortality <-  ifelse(as.numeric(metadata$Mortality) > 1, "NonSurvivor", "Survivor")
```


# 2. filter genes
* non-zero values in at least 20% of samples
* protein coding genes only
```{r}
CUTOFF = round(0.2*ncol(df))

# Filter for genes detected in >=20% of samples
keep <- rowSums(df>0) >= CUTOFF
keep <- names(keep[keep == TRUE])

df_filt <- df[keep,]
```


# FUNCTIONS

### runHierarchicalClustering
```{r}
runHierarchicalClustering <- function(input_df,k_clusters = 3,cluster_id_col_name = "cluster_id") {
  dist_mat <- dist(as.matrix(input_df), method = "euclidean")
  hclusters <- hclust(dist_mat, method = "ward.D2")
  cluster_ids <- cutree(hclusters, k = k_clusters)

  # Convert to dataframe
  cluster_df <- as.data.frame(cluster_ids) %>% setNames(cluster_id_col_name)
  cluster_df$Sample_ID <- row.names(cluster_df)
  cluster_df           <- cluster_df[,c("Sample_ID",cluster_id_col_name)]
  
  cluster_df[,cluster_id_col_name] <- paste("cluster",cluster_df[,cluster_id_col_name],sep = "")
  
  return(cluster_df)
}
```

# 3. Normalize df using voom

### a. model matrix
```{r}
# Make Model
mm <-model.matrix(~ mortality + cohortbatch, data = metadata)
row.names(mm) <- metadata$Sample_ID
```

### b. voom normalization & filter for protein coding genes
```{r}
vwts          <- voom(df_filt, mm, plot = T,normalize.method="quantile")

model_weights <- vwts$weights
df_norm_log   <- as.data.frame(vwts$E)

### Filter for protein coding genes
transcript_rows        <- row.names(df_norm_log)
names(transcript_rows) <- tstrsplit(transcript_rows,"_")[[1]]
transcript_rows_prot   <- transcript_rows[names(transcript_rows) %in% gene_metadata$hgnc_symbol] %>% as.character()

df_norm_log_pc <- df_norm_log[row.names(df_norm_log) %in% transcript_rows_prot,]
```

# 4. Create input for Clustering

### a. input all protein coding genes
```{r}
input_df <- as.data.frame(t(df_norm_log_pc))
```

### b. input - subset the most variable genes
* This could be the topp 1000 for example
```{r}
# Get the most variable genes
gene_var  <- apply(as.matrix(input_df), 2, var)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:1000]

input_df_top_var <- input_df[,names(input_df) %in% top_genes]
```

### c. format variance per gene dataframe
```{r}
gene_var_df <- as.data.frame(gene_var)
gene_var_df$gene_ensembl <- row.names(gene_var_df)
names(gene_var_df)[1] <- "variance"

gene_var_df$top_1000 <- "no"
gene_var_df[gene_var_df$gene_ensembl %in% top_genes,"top_1000"] <- "yes"
gene_var_df$gene <- tstrsplit(gene_var_df$gene_ensembl,"_")[[1]]

# re-order and sort
gene_var_df <- gene_var_df[,c("gene_ensembl","gene","top_1000","variance")]

gene_var_df <- gene_var_df[order(gene_var_df$variance,decreasing = T),]
```

# --------------------

# Hierarchical clustering


# 1. Cluster matrix

* seed setting doesn't really matter here because hierarchical clustering is deterministic and will be the same no matter how many times you redo it

```{r}
set.seed(1234)

# Hierarchical clustering
dist_mat    <- dist(as.matrix(input_df_top_var), method = "euclidean")
hclusters   <- hclust(dist_mat, method = "ward.D2")
cluster_ids <- cutree(hclusters, k = 3) # forms three clusters (k = 3)


# Convert cluster IDs into dataframe
cluster_id_col_name <- "cluster_id_hc_top_var"
cluster_df_top_var <- as.data.frame(cluster_ids) %>% setNames(cluster_id_col_name)
cluster_df_top_var$Sample_ID <- row.names(cluster_df_top_var)
cluster_df_top_var           <- cluster_df_top_var[,c("Sample_ID",cluster_id_col_name)]
cluster_df_top_var[,cluster_id_col_name] <- paste("cluster",cluster_df_top_var[,cluster_id_col_name],sep = "")

```


# 2. Make Dendrogram

### a. make dendrogram object
```{r}
library(ggplot2)

# Calculate the y-axis range for positioning bars below
y_range <- max(dend_data$segments$y)
bar_height <- y_range * 0.05  # Adaptive bar height

dplot_hc <- ggplot() +
  # Main dendrogram
  geom_segment(data = dend_data$segments,
               aes(x = x, y = y, xend = xend, yend = yend),
               linewidth = 0.5) +
  
  # Cluster color bar
  geom_tile(data = dend_data$labels,
            aes(x = x, y = -bar_height, fill = cluster_id_hc_top_var),
            height = bar_height * 2, width = 1) +
  
  # Mortality indicator bar (black for NonSurvivor, white for Survivor)
  geom_tile(data = dend_data$labels,
            aes(x = x, y = -bar_height * 3),
            height = bar_height * 2, width = 1,
            fill = ifelse(dend_data$labels$mortality == "NonSurvivor", "black", "white"),
            color = "gray50", linewidth = 0.1) +
  
  # Add labels for the bars
  annotate("text", x = -5, y = -bar_height, 
           label = "Cluster", hjust = 1.0, size = 3.1, fontface = "bold") +
  annotate("text", x = -5, y = -bar_height * 3, 
           label = "Mortality", hjust = 1.0, size = 3.1, fontface = "bold") +
  
  scale_fill_manual(values = c("cluster1" = "#D55E00",  # Vermillion/Orange
                                "cluster2" = "#009E73",  # Teal/Green
                                "cluster3" = "#0072B2"), # Blue
                    name = "Cluster ID") +
  
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),  # Uncomment this line
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12, face = "bold"),
    plot.margin = margin(5, 5, 5, 40)
  ) +
  labs(x = NULL, y = NULL) +
  coord_cartesian(ylim = c(-bar_height * 5, y_range * 1.05),
                  xlim = c(-8, max(dend_data$labels$x) + 2),
                  clip = "off")

# Add custom legend for mortality
dplot_hc <- dplot_hc +
  guides(fill = guide_legend(order = 1)) +
  annotate("rect", xmin = max(dend_data$labels$x) + 10, 
           xmax = max(dend_data$labels$x) + 12,
           ymin = y_range * 0.15, ymax = y_range * 0.17,
           fill = "black") +
  annotate("rect", xmin = max(dend_data$labels$x) + 10, 
           xmax = max(dend_data$labels$x) + 12,
           ymin = y_range * 0.05, ymax = y_range * 0.07,
           fill = "white", color = "gray50") 

dplot_hc
```

# 3. PDF
```{r}
pdf(file = file.path(plot_dir,"hclust_dendrogram.pdf"),height = 2.0,width = 4.0)
print(dplot_hc)
dev.off()
```


