---
title: "Heirarchical Clustering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# START
```{r}
library(edgeR) 
library(RColorBrewer)
library(gplots)
library(biomaRt) #converting gene names
library(pheatmap)
library(viridis)
library(dplyr)
library(tibble)
library(magrittr)
library(AnnotationDbi)
library(stats4)
library(BiocGenerics)
library(gtools)
library(openxlsx)
library(EnhancedVolcano)
library(stringr)
library(data.table)

library(DESeq2)
library(uwot)
library(gridExtra)
library(ggplot2)
library(cowplot)

library(dendsort)

# directories
projectdirectory      <- "./RNASeq_Analysis"
resouce_dir           <- file.path(projectdirectory,"resources")
input_dir             <- file.path(projectdirectory,"input")
plot_dir              <- file.path(projectdirectory,"plots")
table_dir             <- file.path(projectdirectory,"tables")
object_dir            <- file.path(projectdirectory,"objects")

# Input 
processed_gene_counts_file  <- file.path(input_dir,"input_gene_count_matrix.csv")
metadata_tbm_file           <- file.path(input_dir, "metadata_TBM_formatted.csv")
biomart_metadata_fh         <- file.path(resouce_dir,"biomart_protein_coding_genes.tsv")
```

# --------------------

# FUNCTIONS

### runHierarchicalClustering
```{r}
runHierarchicalClustering <- function(input_df,k_clusters = 3,cluster_id_col_name = "cluster_id") {
  dist_mat <- dist(as.matrix(input_df), method = "euclidean")
  hclusters <- hclust(dist_mat, method = "ward.D2")
  cluster_ids <- cutree(hclusters, k = k_clusters)

  # Convert to dataframe
  cluster_df <- as.data.frame(cluster_ids) %>% setNames(cluster_id_col_name)
  cluster_df$Sample_ID <- row.names(cluster_df)
  cluster_df           <- cluster_df[,c("Sample_ID",cluster_id_col_name)]
  
  cluster_df[,cluster_id_col_name] <- paste("cluster",cluster_df[,cluster_id_col_name],sep = "")
  
  return(cluster_df)
}
```

# --------------------

# 1. Load input data
```{r}
df   <- read.csv(processed_gene_counts_file, row.names = 1,check.names = F)

# biomart protein coding genes
gene_metadata <- read.delim(biomart_metadata_fh,stringsAsFactors = F)

# metadata
metadata           <- read.csv(metadata_tbm_file, stringsAsFactors = FALSE)
row.names(metadata) <- metadata$Sample_ID

metadata <- metadata[metadata$Sample_ID %in% names(df),]
metadata$Mortality <- as.factor(metadata$short_term_Mortality)
metadata$Gender    <- as.factor(metadata$gender)
metadata$Batch     <- as.numeric(metadata$BatchAnalysis)
metadata$cohortbatch   <- paste("TBM",metadata$Batch,sep = "_") %>% as.factor()
metadata$Status        <- metadata$death_shortterm
metadata$Age        <- metadata$age
metadata$TBM_status <- metadata$post_MNGS_TBM_status
metadata$TBM_status[metadata$TBM_status %in% "Definite TBM"] <- "DefiniteTBM"
metadata$TBM_status[metadata$TBM_status %in% "Probable TBM"] <- "ProbableTBM"
metadata$mortality <-  ifelse(as.numeric(metadata$Mortality) > 1, "NonSurvivor", "Survivor")
```


# 2. filter genes
* non-zero values in at least 20% of samples
* protein coding genes only
```{r}
CUTOFF = round(0.2*ncol(df))

# Filter for genes detected in >=20% of samples
keep <- rowSums(df>0) >= CUTOFF
keep <- names(keep[keep == TRUE])

df_filt <- df[keep,]
```


# 3. Normalize df using voom

### a. model matrix
```{r}
# Make Model
mm <-model.matrix(~ mortality + cohortbatch, data = metadata)
row.names(mm) <- metadata$Sample_ID
```

### b. voom normalization
```{r}
vwts          <- voom(df_filt, mm, plot = T,normalize.method="quantile")

model_weights <- vwts$weights
df_norm_log   <- as.data.frame(vwts$E)

### Filter for protein coding genes
transcript_rows        <- row.names(df_norm_log)
names(transcript_rows) <- tstrsplit(transcript_rows,"_")[[1]]
transcript_rows_prot   <- transcript_rows[names(transcript_rows) %in% gene_metadata$hgnc_symbol] %>% as.character()

df_norm_log_pc <- df_norm_log[row.names(df_norm_log) %in% transcript_rows_prot,]
```


# 4. Count non-zero protein coding genes
```{r}
df_bin <- df
df_bin[df_bin > 0] <- 1

transcript_rows <- row.names(df_bin)
names(transcript_rows) <- tstrsplit(transcript_rows,"_")[[1]]
transcript_rows_prot   <- transcript_rows[names(transcript_rows) %in% gene_metadata$hgnc_symbol] %>% as.character()

df_bin_prot <- df_bin[row.names(df_bin) %in% transcript_rows_prot,]
sample_gene_count_df <- colSums(df_bin_prot) %>% as.data.frame()
names(sample_gene_count_df) <- "nonzero_gene_count"
sample_gene_count_df$sample <- row.names(sample_gene_count_df)
sample_gene_count_df$PID    <- tstrsplit(sample_gene_count_df$sample,"_")[[1]]

sample_gene_count_df        <- sample_gene_count_df[,c("sample","PID","nonzero_gene_count")]
```


# 5. Create input for Clustering

### a. input all protein coding genes
```{r}
input_df <- as.data.frame(t(df_norm_log_pc))
```

### b. input - subset the most variable genes
* This could be the topp 1000 for example
```{r}
# Get the most variable genes
gene_var  <- apply(as.matrix(input_df), 2, var)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:1000]

input_df_top_var <- input_df[,names(input_df) %in% top_genes]
```

### c. format into a df for Sumanth
```{r}
gene_var_df <- as.data.frame(gene_var)
gene_var_df$gene_ensembl <- row.names(gene_var_df)
names(gene_var_df)[1] <- "variance"

gene_var_df$top_1000 <- "no"
gene_var_df[gene_var_df$gene_ensembl %in% top_genes,"top_1000"] <- "yes"
gene_var_df$gene <- tstrsplit(gene_var_df$gene_ensembl,"_")[[1]]

# re-order and sort
gene_var_df <- gene_var_df[,c("gene_ensembl","gene","top_1000","variance")]

gene_var_df <- gene_var_df[order(gene_var_df$variance,decreasing = T),]
```


# --------------------

# Hierarchical clustering


# 1. Cluster matrix

### a. all protein coding genes
```{r}
set.seed(1234)

cluster_df_all <- runHierarchicalClustering(input_df,k_clusters = 3,cluster_id_col_name = "cluster_id_hc")
```

### b. top variable genes
```{r}
set.seed(1234)

cluster_df_top_var <- runHierarchicalClustering(input_df_top_var,k_clusters = 3,cluster_id_col_name = "cluster_id_hc_top_var")
```

# 2. Merge all cluster IDs into one table
```{r}
hc_cluster_id_table <- merge(cluster_df_all,cluster_df_top_var,by = "Sample_ID")
```


# 3. Make Dendrogram

### a. redo hclust with top 1000 var genes
```{r}
dist_mat <- dist(as.matrix(input_df_top_var), method = "euclidean")
hclusters <- hclust(dist_mat, method = "ward.D2")
cluster_ids <- cutree(hclusters, k = 3)
```

### b. make dendrogram object
```{r}
dend_data <- ggdendro::dendro_data(hclusters, type = "rectangle")

meta_filt <- metadata[,c("Sample_ID","mortality","TBM_status")]

dend_data$labels <- dend_data$labels %>%
  left_join(hc_cluster_id_table, by = c("label" = "Sample_ID"))
dend_data$labels <- dend_data$labels %>%
  left_join(meta_filt, by = c("label" = "Sample_ID"))
```

### c. plot dendrogram
```{r}
dplot_hc <- ggplot() +
  geom_segment(data = dend_data$segments,
               aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_point(data = dend_data$labels,
             aes(x = x, y = y,
                 color = cluster_id_hc_top_var,
                 shape = mortality),
             size = 3) +
  scale_shape_manual(values = c("Survivor" = 16,   # circle
                                "NonSurvivor" = 15)) + # square
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  labs(color = "ClusterID", shape = "Mortality",
       x = NULL, y = "Height")

dplot_hc
```

### d. PDF
```{r}
plot_dir_main     <- file.path(projectdirectory,"plots_for_figures","main_figures")

pdf(file = file.path(plot_dir_main,"Hclust_plots","hclust_dendrogram.pdf"),height = 5,width = 15)
print(dplot_hc)
dev.off()
```


# --------------------

# Write CSVs

```{r}
write.csv(hc_cluster_id_table,file = file.path(table_dir,"cluster_ID_table_hierarchical_clustering.csv"),row.names = F,quote = F)

# Variable genes
write.csv(gene_var_df,file = file.path(table_dir,"table_gene_variance.csv"),row.names = F,quote = F)
```

