---
title: "Test Cluster stability TBM"
author: "Ravi Dandekar"
date: "2026-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# START
```{r}
library(edgeR) 
library(RColorBrewer)
library(gplots)
library(biomaRt) #converting gene names
library(pheatmap)
library(viridis)
library(dplyr)
library(tibble)
library(magrittr)
library(AnnotationDbi)
library(stats4)
library(BiocGenerics)
library(gtools)
library(openxlsx)
library(EnhancedVolcano)
library(stringr)
library(data.table)

library(DESeq2)
library(uwot)
library(gridExtra)
library(ggplot2)
library(cowplot)

library(dendsort)

# Testing cluster stability
library(M3C)
library(mclust)

# Thresholds

# Color Palettes
COLOR_PAL <- colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(100)
break_1 <- seq(-2,1,0.059)  # first 51 breaks
#break_1 <- seq(-1,1,0.04)  # first 51 breaks

#break_2 <- seq(1,4,0.06)  # last 50 breaks
#break_2 <- break_2[3:length(break_2)]
break_2 <- seq(1,2,0.02)  # last 50 breaks
break_2 <- break_2[2:length(break_2)]
break_list1_4 <- c(break_1,break_2)

COLOR_PAL_RAINBOW <- rainbow(10)

# directories
projectdirectory      <- "/data/rdandekar/rprojects/TBM_RNASeq_Analysis"
resouce_dir           <- file.path(projectdirectory,"resources")
input_dir             <- file.path(projectdirectory,"input")
plot_dir              <- file.path(projectdirectory,"plots")
table_dir             <- file.path(projectdirectory,"tables")
object_dir            <- file.path(projectdirectory,"objects")

# Input 
processed_gene_counts_file  <- file.path(input_dir,"input_gene_count_matrix.csv")
metadata_tbm_file           <- file.path(input_dir, "metadata_TBM_formatted.csv")
biomart_metadata_fh         <- file.path(resouce_dir,"biomart_protein_coding_genes.tsv")
```

# --------------------

# FUNCTIONS

### runHierarchicalClustering
```{r}
runHierarchicalClustering <- function(input_df,k_clusters = 3,cluster_id_col_name = "cluster_id") {
  dist_mat <- dist(as.matrix(input_df), method = "euclidean")
  hclusters <- hclust(dist_mat, method = "ward.D2")
  cluster_ids <- cutree(hclusters, k = k_clusters)

  # Convert to dataframe
  cluster_df <- as.data.frame(cluster_ids) %>% setNames(cluster_id_col_name)
  cluster_df$Sample_ID <- row.names(cluster_df)
  cluster_df           <- cluster_df[,c("Sample_ID",cluster_id_col_name)]
  
  cluster_df[,cluster_id_col_name] <- paste("cluster",cluster_df[,cluster_id_col_name],sep = "")
  
  return(cluster_df)
}
```

# --------------------

# 1. Load input data

### a. load input dataframe and metadata
```{r}
df   <- read.csv(processed_gene_counts_file, row.names = 1,check.names = F)

# biomart protein coding genes
gene_metadata <- read.delim(biomart_metadata_fh,stringsAsFactors = F)

# metadata
metadata           <- read.csv(metadata_tbm_file, stringsAsFactors = FALSE)
row.names(metadata) <- metadata$Sample_ID

metadata <- metadata[metadata$Sample_ID %in% names(df),]
metadata$Mortality <- as.factor(metadata$short_term_Mortality)
metadata$Gender    <- as.factor(metadata$gender)
metadata$Batch     <- as.numeric(metadata$BatchAnalysis)
metadata$cohortbatch   <- paste("TBM",metadata$Batch,sep = "_") %>% as.factor()
metadata$Status        <- metadata$death_shortterm
metadata$Age        <- metadata$age
metadata$TBM_status <- metadata$post_MNGS_TBM_status
metadata$TBM_status[metadata$TBM_status %in% "Definite TBM"] <- "DefiniteTBM"
metadata$TBM_status[metadata$TBM_status %in% "Probable TBM"] <- "ProbableTBM"
metadata$mortality <-  ifelse(as.numeric(metadata$Mortality) > 1, "NonSurvivor", "Survivor")
```

### b. load cluster IDs and top var genes
```{r}
df_h_cluster_ids <- read.csv(file.path(table_dir,"cluster_ID_table_hierarchical_clustering_Oct2025.csv"),stringsAsFactors = F)
df_top_var_genes <- read.csv(file = file.path(table_dir,"table_gene_variance_in_final_cohort.csv"),stringsAsFactors = F)
```


# 2. filter genes
* non-zero values in at least 20% of samples
* protein coding genes only
```{r}
CUTOFF = round(0.2*ncol(df))

# Filter for genes detected in >=20% of samples
keep <- rowSums(df>0) >= CUTOFF
keep <- names(keep[keep == TRUE])

df_filt <- df[keep,]
```


# 3. Normalize df using voom

### a. model matrix
```{r}
# Make Model
mm <-model.matrix(~ mortality + cohortbatch, data = metadata)

# Change Dead vs Alive to Alive vs dead (Alive = 1, Dead = 0)
#comp_col <- mm[,2]
#comp_col[comp_col == 1] <- 2
#comp_col[comp_col == 0] <- 1
#comp_col[comp_col == 2] <- 0
#mm[,2] <- comp_col

row.names(mm) <- metadata$Sample_ID
```

### b. voom normalization
```{r}
vwts          <- voom(df_filt, mm, plot = T,normalize.method="quantile")

model_weights <- vwts$weights
df_norm_log   <- as.data.frame(vwts$E)

### Filter for protein coding genes
transcript_rows        <- row.names(df_norm_log)
names(transcript_rows) <- tstrsplit(transcript_rows,"_")[[1]]
transcript_rows_prot   <- transcript_rows[names(transcript_rows) %in% gene_metadata$hgnc_symbol] %>% as.character()

df_norm_log_pc <- df_norm_log[row.names(df_norm_log) %in% transcript_rows_prot,]
```


# 4. Count non-zero protein coding genes
```{r}
df_bin <- df
df_bin[df_bin > 0] <- 1

transcript_rows <- row.names(df_bin)
names(transcript_rows) <- tstrsplit(transcript_rows,"_")[[1]]
transcript_rows_prot   <- transcript_rows[names(transcript_rows) %in% gene_metadata$hgnc_symbol] %>% as.character()

df_bin_prot <- df_bin[row.names(df_bin) %in% transcript_rows_prot,]
sample_gene_count_df <- colSums(df_bin_prot) %>% as.data.frame()
names(sample_gene_count_df) <- "nonzero_gene_count"
sample_gene_count_df$sample <- row.names(sample_gene_count_df)
sample_gene_count_df$PID    <- tstrsplit(sample_gene_count_df$sample,"_")[[1]]

sample_gene_count_df        <- sample_gene_count_df[,c("sample","PID","nonzero_gene_count")]
```


# 5. Create input for Clustering

### a. input all protein coding genes
```{r}
input_df <- as.data.frame(t(df_norm_log_pc))
```

### b. input - subset the most variable genes
* This could be the topp 1000 for example
```{r}
# Get the most variable genes
top_genes <- df_top_var_genes[df_top_var_genes$top_1000 %in% "yes",]
top_genes <- top_genes$gene_ensembl

input_df_top_var <- input_df[,names(input_df) %in% top_genes]
```

# --------------------

# Cluster stability

# 1. make inputs
```{r}
# 1) Make the matrix M3C expects: rows = genes/features, columns = samples
x_counts <- as.matrix(input_df_top_var)
x <- t(x_counts)

# change order in 
row.names(df_h_cluster_ids) <- df_h_cluster_ids$Sample_ID
df_h_cluster_ids            <- df_h_cluster_ids[colnames(x),]

# Make safe sample IDs once, then use them everywhere
orig_ids <- colnames(x)
safe_ids <- make.names(orig_ids, unique = TRUE)

colnames(x) <- safe_ids

des <- data.frame(
  ID = safe_ids,
  hc_id = df_h_cluster_ids$cluster_id_hc_top_var[match(orig_ids, df_h_cluster_ids$Sample_ID)]
)
#row.names(des) <- des$ID
```


# 2. M3C evaluation
* clusteralg = "hc"  - used to consistency of how our clusters were assembled
* clusteralg = "pam" - default method for M3C consensus clustering
```{r}
stopifnot(identical(colnames(x), des$ID))
stopifnot(nrow(des) == ncol(x))

# 5) Run M3C with hierarchical clustering
#debugonce(M3Creal)
res <- M3C(
  mydata     = as.data.frame(x),
  des        = des,
  clusteralg = "hc",
  method     = 1,       # Monte Carlo null model
  maxK       = 6,       # must be >= 3 for your case
  iters      = 25,
  repsref    = 100,
  repsreal   = 100,
  pItem      = 0.8,
  objective  = "entropy",
  seed       = 123,
  removeplots = FALSE,
  silent     = FALSE
)

# 6) Inspect overall K-level quality statistics
res$scores
```

# 3. Extract results
```{r}
k <- 3

# K-specific M3C outputs
ann3  <- res$realdataresults[[k]]$ordered_annotation
cc3   <- res$realdataresults[[k]]$consensus_matrix
data3 <- res$realdataresults[[k]]$ordered_data

# Match your provided cluster labels to the M3C-ordered samples
cmp3 <- data.frame(
  sample_id   = rownames(ann3),
  des_cluster = des$hc_id[match(rownames(ann3), des$ID)],
  m3c_cluster  = as.character(ann3$consensuscluster),
  stringsAsFactors = FALSE
)

# Overall agreement between your labels and M3C's K=3 consensus labels
tab3 <- table(cmp3$des_cluster, cmp3$m3c_cluster)
ari3 <- adjustedRandIndex(cmp3$des_cluster, cmp3$m3c_cluster)

tab3
ari3

# How each of your clusters aligns to the M3C clusters
cluster_purity <- apply(tab3, 1, function(z) max(z) / sum(z))
cluster_purity

# Simple chi-squared test of association
chisq.test(tab3)
```

